Estimating Daily Bicycle Counts in Seattle from Seasonal and Weather Factors

### 2016-02-07 article/article.tex
edits based on reviewer comments

Bicycle counts were collected at this location continuously by the
City of Seattle using an in-sidewalk counter manufactured by
EcoCounter. When a bicycle passes over an induction loop embedded in
the sidewalk on either side of the Fremont Bridge, the counter
registers the bicycle. Bicyclists may legally choose to ride in the
roadway instead of the sidewalk, and would thus not be detected by the
counter. However, we believe these crossings are rare at this location
due to the design of the facility, which directs bicyclists to enter
the sidewalk, and from our own experience riding and observing other
riders. Counts are aggregated into 15 minute intervals by the City of
Seattle, and are made available to the public via the City of
Seattle's data
portal \parencite{City-of-Seattle:aa,City-of-Seattle:ab}.

@@

We downloaded and processed these data programmatically using the R
programming language along with several add-on
packages \parencite{Grolemund:2011aa,Wickham:2011aa,Couture-Beil:2014aa,Lang:2014aa,R-Core-Team:2014aa}.\footnote{In
addition, the code used to run the analysis, as well as the final
dataset are available as a github repository: \url{https://github.com/pschmied/bikecounts-mle}}
Bicycle counts were aggregated by day, and then joined to weather data
by date. In addition to the variables collected from these two
sources, we were also interested in controlling for holidays and
whether or not the nearby University of Washington was in session.
These data were collected and coded manually from the University of
Washington's historic academic calendars.

@@

We deemed the University of Washington variable important in part
because of the Fremont Bridge's proximity and connection via the Burke
Gilman Trail to the University of Washington. We also felt that this
variable was a suitable proxy for the ``school season,'' which more
broadly captures whether or not other local schools are in session.
The academic calendars of the various local schools do not align
perfectly, however they still overlap substantially with the
University of Washington, which is itself the largest educational
institution in the region.

Inclusion of the holiday variable was an attempt to account for
some low outlier counts. Upon inspection of the dataset, Christmas and
Thanksgiving in particular had very low counts of bicycles relative to
the days preceding and following. Relatedly, but not accounted for by
any variable in our model, are some of the high outlier counts. Upon
inspection, some of the highest counts were observed on National Bike
to Work Day and on the day of the Fremont Solstice Parade, which
typically draws large numbers of bicyclists as participants and
spectators. The omission of such a variable is justified based on the
few occurrences of high outlier counts, and our desire for this model
to only include variables that could be collected or straightforwardly
adapted to other locations.

@@

Maximum precipitation, which measures the maximum inches of
precipitation that occurred in any hour throughout the day, was chosen
rather than average precipitation based on the notion that bicyclists
might make travel decisions based on a likely worst case scenario.
This assumption is slightly more problematic than our assumptions
about temperature, in that we do expect bicyclists to be at least
somewhat sensitive to average conditions or conditions observed at
their time of departure. As in the case of temperature, this
simplification would be less justifiable in locations that experience
greater daily variation in precipitation or in locations that have a
predictable pattern of precipitation during certain hours.

@@

Figure \ref{fg:seasonality} shows that we see a substantial increase
in bicycle volumes when the University of Washington was in session.
With all other factors held constant we see that, on days when the
university is in session, we would expect an average of approximately
367 additional bicycles observed. Similarly, we see a roughly linear
increase in bicycles with increased day length.

@@

Temperature, in contrast to precipitation, has a clearly positive
association with increased numbers of bicyclists. Temperature squared
was not significant in our model at the $p < 0.05$ level. We were
somewhat surprised to not see evidence for a leveling off in counts at
very high temperatures---likely due to Seattle's moderate summer
climate, and the lack of extremely warm days in our dataset.

@@

One variable missing from our analysis that we would have liked to
include is cloud cover. This variable would have allowed us to account
for the possibility of brilliant, sunny, but cold winter days, as well
as warm, dry, but gloomy summer days. Unfortunately, the dataset
downloaded from Forecast.io included a great deal of missing values
for this variable, and thus future work might include identification
of a more suitable source of these data.

>>>>

Bicycle counts were collected at this location continuously by the
City of Seattle using an in-sidewalk counter manufactured by
EcoCounter. When a bicycle passes over an induction loop embedded in
the sidewalk on either side of the Fremont Bridge, the counter
registers the bicycle. Bicyclists may legally choose to ride in the
roadway instead of the sidewalk, and would thus not be detected by the
counter. However, we believe these crossings are rare at this location
due to the design of the facility (metal roadway grating and wide
sidewalks on the approach), which directs bicyclists to enter the
sidewalk, and from our own experience riding and observing other
riders. Counts are aggregated into 15 minute intervals by the City of
Seattle, and are made available to the public via the City of
Seattle's data
portal \parencite{City-of-Seattle:aa,City-of-Seattle:ab}.

@@

We downloaded and processed these data programmatically using the R
programming language along with several add-on
packages \parencite{Grolemund:2011aa,Wickham:2011aa,Couture-Beil:2014aa,Lang:2014aa,R-Core-Team:2014aa}. In
addition, the code used to run the analysis, as well as the final
dataset are available as a github repository [NEED IN-TEXT CITATION FOR https://github.com/pschmied/bikecounts-mle]
Bicycle counts were aggregated by day, and then joined to weather data
by date. In addition to the variables collected from these two
sources, we were also interested in controlling for holidays and
whether or not the nearby University of Washington was in session.
These data were collected and coded manually from the University of
Washington's historic academic calendars.

@@

We deemed the University of Washington variable important in part
because of the Fremont Bridge's proximity and connection via the Burke
Gilman Trail to the University of Washington. We also felt that this
variable was a suitable proxy for the ``school season,'' (i.e. times
when significant numbers of students are not on holiday) and more
broadly captures whether or not other local schools are in session.
The academic calendars of the various local schools do not align
perfectly, however they still overlap substantially with the
University of Washington, which is itself the largest educational
institution in the region.

Inclusion of the holiday variable (as recognized by the University of
Washington) was an attempt to account for some low outlier
counts. Upon inspection of the dataset, Christmas and Thanksgiving in
particular had very low counts of bicycles relative to the days
preceding and following. Relatedly, but not accounted for by any
variable in our model, are some of the high outlier counts. Upon
inspection, some of the highest counts were observed on National Bike
to Work Day and on the day of the Fremont Solstice Parade, which
typically draws large numbers of bicyclists as participants and
spectators. The omission of such a variable is justified based on the
few occurrences of high outlier counts, and our desire for this model
to only include variables that could be collected or straightforwardly
adapted to other locations.

@@

Maximum precipitation, which measures the maximum inches of
precipitation that occurred in any hour throughout the day, was chosen
rather than average precipitation based on the notion that bicyclists
might make travel decisions based on a likely worst case scenario.
This assumption is slightly more problematic than our assumptions
about temperature, in that we do expect bicyclists to be at least
somewhat sensitive to average conditions or conditions observed at
their time of departure. As in the case of temperature, this
simplification would be less justifiable in locations that experience
greater daily variation in precipitation, greater variety of
precipitation types (e.g. snow), or in locations that have a
predictable pattern of precipitation during certain hours.

@@

Figure \ref{fg:seasonality} shows that we see a substantial increase
in bicycle volumes when the University of Washington was in session.
With all other factors held constant we see that, on days when the
university is in session, we would expect to observe an average of
approximately 367 additional bicycles as compared to days when the
university is not in session. Similarly, we see a roughly linear
increase in bicycles with increased day length.

@@

Temperature, in contrast to precipitation, has a clearly positive
association with increased numbers of bicyclists. Temperature squared
was not significant in our model at the $p < 0.05$ level. We were
somewhat surprised to not see evidence for a leveling off in counts at
very high temperatures---likely due to Seattle's moderate summer
climate, and the lack of extremely warm days in our dataset (85
degrees Fahrenheit was the maximum observed).

@@

One variable missing from our analysis that we would have liked to
include is cloud cover. This variable would have allowed us to account
for the possibility of brilliant, sunny, but cold winter days, as well
as warm, dry, but gloomy summer days. On such days, we might observe
counts otherwise uncharacteristic of the season. Unfortunately, the
dataset downloaded from Forecast.io included a great deal of missing
values for this variable, and thus future work might include
identification of a more suitable source of these data.
